---
title: "An Analysis of Glassdoor Data Analyst Listings"
output:
  html_document:
    toc: true
    toc_float: true
---

&nbsp;  

## Using data cleaning, ggplot and dplyr to analyze numerically and visually data analyst positions by location, industry, and salary

&nbsp;  

This projects data was collected using webscraping from Glassdoor to compile data analyst positions listed across the united states.


I am interested in finding out:

* The distribution of salaries for data analyst positions
* What different industries are hiring data analysts?
* What these industries offer data analysts in salaries?
* Where are most of these positions located for analyst positions, by their respective industry?

&nbsp;  

I'll be using visualizations of the data to help answer these questions, as well as the output of the data itself

### **1. Loading in and cleaning the data**

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(readr)
library(dplyr)
library(plotly)
library(ggplot2)
```

```{r, echo = FALSE, warning=FALSE}
data <- read_excel("C:/Users/pmcgee/Downloads/Analyst Portfolio/datasets/analyst_positions_data.xlsx")
glimpse(data)
```

&nbsp;  

Defintely will have some data cleaning to do with this. I'm going to:

* remove extra text from salary column
* remove the rating that shouldn't be in the company name column
* remove rows where salary = ""
* make salary and rating a numeric data type for analysis
```{r, include=FALSE}
data <- data %>% rename(Salary = `Salary Estimate` )
data_2 <- data

data_2$Salary <- gsub("[(Glassdoor est.)K$]", "", data$Salary)
unique(data_2$Salary)
data_3 <- data_2

data_3$Salary <- sub("\\-.*",'', perl = T, data_3$Salary)
unique(data_3$Salary)

#Remove rating from company name
data_3 <- data_3 %>% rename(Comp_name = `Company Name` )
data_3$Comp_name <- sub("\\s+[^ ]+$", "", data_3$Comp_name)


#remove rows where salary = ""
data_3 <- subset(data_3, Salary!="")
unique(data_3$Salary)

#make salary and rating numeric
data_3$Salary <- as.numeric(data_3$Salary)
data_3$Rating <- as.numeric(data_3$Rating)
```

```{r, echo = FALSE}
head(data_3)
```
I chose to keep the smallest value in the estimated salary range, so our data may look like salaries are on the low side.

&nbsp;  

### **2. Visualizing distribution of salaries**
I'll create a histogram and print summary statistics to see what the salary range is

```{r, echo = FALSE}
summary(data_3$Salary)
hist_sal<- data_3 %>% ggplot(aes(Salary)) + geom_histogram(binwidth = 2, color = "white", fill = "black") +
  labs(title = "Data Analyst Positions Salaries")
hist_sal
```

&nbsp;  

### **3. What industries are looking for Analysts?**
I'll see which industries have more analyst position postings right now.
```{r, echo = FALSE}
data_4 <- data_3
indus_count <- data_4 %>% count(Industry) %>% arrange(desc(n))
indus_count <- subset(indus_count, Industry!="-1") 
head(indus_count,20)
```

&nbsp;  

Now I'll shorten the text in the industry column to just keep the first word and create a plot to visualize the results of industries with 40 or more job postings.
```{r, echo =FALSE}
indus_count <- indus_count %>% filter(n >= 40)
indus_count$Industry <- gsub(' [A-z ]*', '' ,indus_count$Industry)
indus_n <- indus_count

#shortening names
indus_n$Industry <- gsub("&", "", indus_n$Industry)
indus_n

#Plots of most frequent industries looking for analysts

#all black theme
blackout <- theme(
  # get rid of panel grids
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  # Change plot and panel background
  plot.background=element_rect(fill = "gray"),
  panel.background = element_rect(fill = 'black'),
  # Change legend 
  legend.direction = "horizontal",
  legend.background = element_rect(fill = "black", color = NA),
  legend.key = element_rect(color = "gray", fill = "black"),
  legend.title = element_text(color = "white"),
  legend.text = element_text(color = "white")
)

indus_plot <- indus_n %>% ggplot(aes(x = Industry, y = n, fill = Industry)) + geom_col() +
    theme(axis.text.x = element_text(angle = 90), legend.position="none") + ylab("Number of Positions") + blackout
```
```{r, echo=FALSE,warning=FALSE}
indus_plot
```

&nbsp;  

### **4. What do data anlysts get paid in different industries?**

Let's see how much data analysts get paid on average based on what industry they work in.

I'll use the top 20 industries by amount of jobs postings on Glassdoor to determine this.
```{r, echo=FALSE}
Sal_Indus <- data_4 %>% select(Salary, Comp_name, Industry, Location)
Sal_Indus <- subset(Sal_Indus, Industry!="-1")
Sal_Indus$Industry <- gsub(' [A-z ]*', '' ,Sal_Indus$Industry)
Sal_Indus$Industry <- gsub("&", "", Sal_Indus$Industry)
Sal_Indus$Industry <- gsub(",", "", Sal_Indus$Industry)

counts <- Sal_Indus %>% count(Industry) %>% arrange(desc(n))
#counts
top20 <- head(counts, 20)
top20
names <- top20 %>% select(Industry)
#class(names)
names <- as.vector(names)
#class(names)
avector <- names[['Industry']]
```

These are what industries pay data analysts the most on average
```{r, echo = FALSE}
Industry_avg_sal <- Sal_Indus %>%
  group_by(Industry) %>%
    summarise(avg_sal = mean(Salary)) %>% 
      arrange(desc(avg_sal))
Industry_avg_sal

Industry_avg_sal_2 <- head(Industry_avg_sal,10)
Industry_avg_sal_plot <- Industry_avg_sal_2 %>% ggplot(aes(x = Industry, y = avg_sal, fill = Industry)) +
  geom_col() + 
    labs(title = "Average Salary for all Industries", y = "Average Salary") +
      theme(axis.text.x = element_text(angle = 90), legend.position="none",  axis.title.x = element_blank()) 
Industry_avg_sal_plot

```

Now. let's see how much the industries with the most amount of jobs posted pay data analysts.
```{r, echo=FALSE}
Highest_sal <- Industry_avg_sal[Industry_avg_sal$Industry %in% avector, ]
Sal_plot <- Highest_sal %>% ggplot(aes(x = reorder(Industry, - avg_sal), y = avg_sal, fill = Industry)) +
  geom_col() + theme(axis.text.x = element_text(angle = 90),legend.position="none",  axis.title.x = element_blank())  + 
  labs(title = "Average Salary for Data Analysts by Industry with most Postings", x = "Average Salary")

Sal_plot
```

&nbsp;  

### **5. What location has the most jobs for each industry**

Now I'll investigate to find out what industry in what cities have the most amount of jobs.My guess would be that New YOrk would have the most positions and the most industries that are currently hiring.

```{r, echo = FALSE}

#Where are most jobs located by industry
data_5 <- data_4
data_5 <- data_5 %>% select(Salary, Comp_name, Industry, Location)
#shorten industry names
data_5$Industry <- gsub(' [A-z ]*', '' , data_5$Industry)
data_5$Industry <- gsub("&", "", data_5$Industry)
data_5$Industry <- gsub(",", "", data_5$Industry)
#droppping "-1" in Industry
data_5 <- data_5[!grepl("-1", data_5$Industry), ]
#Drop state after comma
data_5$Location <- gsub(",.*","",data_5$Location)

loc_count <- data_5 %>% group_by(Industry) %>% count(Location) %>% arrange(desc(n))
loc_count
loc_count <- loc_count[1:20,]
```

&nbsp;  

New York appears to have the most postings. To make this list easier to comprehend, I'll plot number of positions per industry and facet wrap by city.
```{r,echo=FALSE}
loc_count_plot <- loc_count %>% ggplot(aes(x = Industry, y = n, fill = Industry)) + geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) + 
    facet_wrap(~ Location) + theme(axis.title.x = element_blank(),axis.title.y = element_blank()) +
      ggtitle("Number of Analyst Jobs per Industry by Location") +
        theme(plot.title = element_text(hjust = 0.5))

loc_count_plot
```

&nbsp;  

### **6. Conclusion**

New York is the city with the most jobs in the most industries that are hiring right now.

Biotech is the industry which pays the most on average out of the top 20 industries with the most postings, while News pays the most overall on average.

The top industries hiring include IT, Staffing, Health Care, Computer Hardware, Consulting and many more.

A pessimistic estimate of the range of salaries for data analysts are between $29,000 -$112,000.

&nbsp;  


