---
title: "Insurance Cost Prediction with Linear Regression"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyverse)
library(readxl)
library(readr)
library(dplyr)
library(plotly)
library(ggplot2)
library(GGally)
library(psych)
library(relaimpo)
```

&nbsp;  

#### Using linear regression, exploratory data analysis and data visualization to predict the amount which individuals have spent on health care.

This project uses the medical cost personal dataset to see what health variables are most impactful regarding the amount an individual has spent on medical costs and how much those costs are.

variables include age, sex, bmi, # children, smoker, region of the us the person is from and the amount they spent on medical bills.

### **1. Reading in the data**
```{r, include=FALSE}
dat <- read_excel("C:/Users/pmcgee/Downloads/Analyst Portfolio/datasets/insurance.xlsx")
```
```{r}
head(dat)
str(dat)
```
The variables in this dataset include age, sex, bmi, # of children, region, smoker and the amount charged.

&nbsp;  

### **2. Exploratory data analysis**
I'll have a look at a quick summary of the data
```{r,echo=FALSE}
summary(dat)
```
&nbsp;  

To see what variables may most closely impact an individuals medical cost, I'll create a correlation matrix of the numeric variables in this data set.
```{r, echo=FALSE}
cor_dat <- dat %>% dplyr::select(age, bmi, children, charges)
cor_matrix <- cor(cor_dat)
cor_matrix
```

Age is most closely correlated with spending on ones health.

I'll view these correlations visually.
```{r, echo = FALSE}
pairs.panels(dat[c("age", "bmi", "children", "charges")])
```

Once again these plots show that age is most closely correlated with charges among the numeric variables. None of the numeric variables are highly correlated between one another, so this eliminates the problem of multicollinearity. A final observation from these plots is to note that age and charges may not have a linear relationship.

&nbsp;  

#### **Plotting the non-numeric variables**

Plotting the charges by region
```{r, echo=FALSE}
region_plot <- dat %>% ggplot(aes(x = region, y = charges, fill = region)) + geom_boxplot() + 
  ggtitle(("Boxplot of Medical Charges Per region"))
region_plot
```

Those in southeast have the highest third quartile in spending, with some outliers spending the most out of all individuals.

&nbsp;  

Plotting charges by gender
```{r, echo=FALSE}
sex_plot <- dat %>% ggplot(aes(x = sex, y = charges, fill = sex)) + geom_boxplot() + 
  ggtitle(("Boxplot of Medical Charges by Sex"))
sex_plot
```

Men have a higher third quartile than spending than females.

&nbsp;  

Plotting charges for smokers and non-smokers
```{r, echo=FALSE}
smoke_plot <- dat %>% ggplot(aes(x = smoker, y = charges, fill = smoker)) + geom_boxplot() + 
  ggtitle(("Boxplot of Medical Charges for Smoker/non-Smoker"))
smoke_plot
```

smokers spend significantly more than non-smokers

&nbsp;  

Plotting charges by number of children
```{r, echo=FALSE}
kid_plot <- dat %>% ggplot(aes(x = as.factor(children), y = charges)) + geom_boxplot(fill = c(2:7)) + 
  ggtitle(("Boxplot of Medical Charges Per # of Children"))
kid_plot
```

Those with 3 and 4 children spend the most on average, while those with 0 children also spend a fair amount. It is likely that there are less data points for those with more children, since it is rarer to have many children. 

I'll investigate this clam
```{r, warning=FALSE, message=FALSE}
kid_spend <- dat %>% group_by(children) %>% summarize(money_spent = sum(charges), num_fam = n())
kid_spend
```

There are less data points for families with less kids, because fewer families tend to have more than 3 kids.

&nbsp;  

### **3. Creating a new variable for the model**
I'll create a new variable for those who are labeled as obese based on their BMI. According to the CDC, if your BMI is above 30 you are considered obese. Obese will be the name of the new variable, with responses being 'yes' or 'no.'
```{r, include=FALSE}
dat_2 <- dat
```
```{r}
dat_2$obese <- ifelse(dat_2$bmi> 30,"yes" ,"no")
head(dat_2)
```

&nbsp;  

I'll create a plot to show the impact of being obese on medical expenses.
```{r, echo=FALSE}
obese_plot <- dat_2 %>% ggplot(aes(x = obese, y = charges, fill = obese)) + geom_boxplot() +
  ggtitle("Boxplot for Medical Exp. by Obesity")
obese_plot
```

Charges are slightly higher on average, with many more extremes in charges for those who are obese.

&nbsp;  

Intuition tells us obese smokers are usually at the most risk of having health problems
I'll make a plot for this demographic.
```{r, include=FALSE}
dat_3 <- dat_2
class(dat_3$smoker)
dat_3$smoker<- as.factor(dat_3$smoker)
levels(dat_3$smoker)<- c("Non-Smoker", "Smoker")
```
```{r,echo=FALSE}
fat_smoker <- dat_3 %>% ggplot(aes(x = obese, y = charges, fill = obese)) + geom_boxplot() +
  facet_wrap(~ smoker)
fat_smoker
```

Obese Smokers are at by far the most risk, while smokers who are not obese are at more risk than those who are just obese.

&nbsp;  

### **4. Model Creation**
I'll create a few models so that I can tune the variables selected to maximize my R-squared and adjusted R-squared, which  represents the proportion of the variance for a dependent variable that's explained by an independent variable or variables in a regression model.

First I will implement a kitchen sink approach using the variables.
```{r}
model_1 <- lm(charges~ ., dat_2)
summary(model_1)
```
The variables which we see have a significant impact in determing the amount of personal will be charged for medical costs at a p-value of 0.05 include age, bmi, children, obeseyes, and regionsouthwest.

&nbsp;  

Now, a new model using only significant variables
```{r}
model_2 <- lm(charges ~ age + bmi + children + smoker + region + obese, data = dat_2)
summary(model_2)
```
I removed the variables that are less important, but R-squared has barely improved.

&nbsp;  

I'll create another new variable where I square every individuals age and then create an interaction variable in the model to show the impact of being both a smoker and being obese had an a persons medical cost.

Squaring the individuals age accounts for non-linear relationship between the age and charges variables which was shown in the graph in the exploratory data analysis. Both these change to the model should increase the R-squared.

```{r, include=FALSE}
dat_2 <- dat_2 %>% mutate(age_sq = age^2)
head(dat_2)
```
```{r,echo=FALSE}
model_4 <- lm(charges ~ age + age_sq + bmi + children + region + smoker * obese, data = dat_2)
summary(model_4)
```
The R-squared of this model had increased to 0.86 from 0.75 in the previous models.
Looking at the estimates of this model, we can see some interesting was which our independent variables interact with the dependent variable.

1. All other things equal, those in the southwest tend to pay $1,161.00 less for medical costs than if they are from the north east.

2. All other things equal, as a persons body mass index increases by 1 point, they pay an additional $113.00.

3. All other things equal, smokers pay an additional $13,401.00.

4. All other things equal, those labeled as obese actually pay $939.00 less than those who are not obese. BMI, as well as obesity can be a bit controversial as determining a persons overall health. It is definitely possible for someone categorized as obese based on their BMI to be perfectly healthy.

5. All other things equal, obese smokers pay by far the most in additional medical costs at $19,866.00.

&nbsp;  

### **5. Using the linear regression model to predict medical costs**
 I will create predictions for those who are in the insurance dataset using the most recent linear regression model.
 ```{r}
 pred_charge <- predict(model_4, data = dat_2)
head(pred_charge)
```
Now I'll create 4 new people to predict what their medical costs will be.

1. **Bob**: 19 years old, BMI 27.9, has no children, smokes, from northwest region.

2. **Lisa**: 40 years old, BMI 50, 2 children, doesn't smoke, from southeast region.

3. **John**: 30 years old. BMI 31.2, no children, doesn't smoke, from northeast region.

4. **Amy**: 21 years old. BMI 22, no children, doesn't smoke, from southwest region.

```{r new_test, message=FALSE, warning=FALSE, paged.print=TRUE}
Bob <- data.frame(age = 19,
                  bmi = 27.9,
                  age_sq = 361,
                  children = 0,
                  obese = "no",
                  smoker = "yes",
                  region = "northwest")
print(paste0("Health care charges for Bob: $", round(predict(model_4, Bob), 2)))


Lisa <- data.frame(age = 40,
                   bmi = 50,
                   children = 2,
                   smoker = "no",
                   obese = 'yes',
                   age_sq = 1600,
                   region = "southeast")
print(paste0("Health care charges for Lisa: $", round(predict(model_4, Lisa), 2)))

John <- data.frame(age = 30,
                   bmi = 31.2,
                   children = 0,
                   smoker = "yes",
                   obese = 'yes',
                   age_sq = 900,
                   region = "northeast")
print(paste0("Health care charges for John: $", round(predict(model_4, John), 2)))


Amy <- data.frame(age = 21,
                   bmi = 22,
                   children = 0,
                   smoker = "no",
                   obese = 'no',
                   age_sq = 441,
                   region = "southwest")
print(paste0("Health care charges for Amy: $", round(predict(model_4, Amy), 2)))
```

### **6. Conclusions**

The variables in our model which are most significant in predicting the amount one has spent on medical bills include 

> a persons age squared
> their bmi
> the number of children they have
> if they are from the southeast or southwest
> if they are a smoker
> if they are obese
> if they are an obese smoker

Being obese, from the southwest or southeast actually have a negative impact on a persons medical bills while the rest all increase the amount spent.

Models can be tuned an improved upon by building several to see what variables may or may not be significant. They can also be improved by adding interaction terms to such how multiple variables together may impact the target variables.

This model was able to be used to predict outside individuals amount they likely spent on medical expenses. This models accuracy could perhaps be improved by splitting it into a training and test dataset to more accurately view how correct the predictions were within the data we have on hand.

This data would likely also be a good choice to implement a machine learning model on, but that is for another day.

&nbsp;  
