---
title: "Scouting an Athletics Team"
author: "Paul McGee"
date: "8/17/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using the dplyr package, data cleaning, functions and more to draft and compete a team in a javelin event.

This is a project that I did using  track and field data using R to show which athletes I would want on my team, fantasy draft style. The event of interest I picked was the women's javelin.

&nbsp;  

### **1. Reading in the data.**

```{r, include = FALSE}
library(tidyverse)
library(readxl)
data <- read_excel("C:/Users/pmcgee/Downloads/Analyst Portfolio/datasets/athletics.xlsx")
```
```{r}
head(data)
```

#### This is what first 5 rows of the uncleaned data looks like.

&nbsp;  
I'm going to clean it up a bit. The changes I'll make include: 

1. Male and Female are their own columns

2. Going to gather the data from wide to long so there is only 1 column for throws

3. Select the Womens Javelin event

4. Change the column labeled "Distance" to the data type to numeric to make later analysis easier 


```{r, include = FALSE}
# Select the results of interest: women's javelin
head(data)
javelin <- data %>% filter(Event == "Javelin", Male_Female == "Female")
javelin$Event <- NULL
javelin$Male_Female <- NULL
head(javelin)

# Assign the tidy data to javelin_long
head(javelin)
lapply(javelin,class)
javelin_long <- javelin %>% gather(Flight, Distance, Flight1:Flight6)

# Make Flight a numeric
javelin_long$Flight <- gsub("Flight", "", javelin_long$Flight)
javelin_long$Flight <- as.numeric(javelin_long$Flight)
```

```{r, echo = FALSE }
# Examine the first 6 rows
head(javelin_long,2)
```
&nbsp;  

### **2. Creating sumary statistics and creating new columns** 

The next step is to create the total results for each event for each athlete, as well as create some summary statistics to better describe their performance.

These stats include total distance thrown, standard deviation of their throws and the number of successful throws they had.

```{r}
javelin_totals <- javelin_long %>%
  filter(Distance > 0) %>%
    group_by(Athlete, EventID) %>%
      summarize(TotalDistance = sum(Distance), 
          StandardDev = round(sd(Distance),3),
              Success = n())
```
```{r, include = FALSE}
javelin_totals[1:5,]
```

&nbsp;  

#### Now I'll find the difference between the persons last 3 and first 3 throws and then use a left join to combine this data with the javelin_total data frame.

```{r, results = 'hide'}
# Create first_3, last_3, and diff between the two
javelin <- javelin %>% mutate(first_3 = Flight1 + Flight2 + Flight3 , last_3 = Flight4 + Flight5 + Flight6, diff = last_3 - first_3)

#Select columns to to keep from javelin and javelin totals
keep <- c("Athlete", "TotalDistance", "StandardDev", "Success", "diff")
javelin_totals <- left_join(javelin_totals, javelin, by = c("EventID", "Athlete"), all.x = T) %>% subset( select = keep)
head(javelin_totals, 5)
```

&nbsp;  

### **3. Normalizing the data for easier comparison**

I'll normalize the summary statistics so they can be properly compared. 
The goal of normalization is to change the values of numeric columns in the dataset to a common scale, without distorting differences in the ranges of values so that they can be used for comparison. 
```{r, results="hide", warning=FALSE}
#function to normalize summary stats

norm <- function(result) {
    (result - min(result)) / (max(result) - min(result))
}

aggstats <- c("TotalDistance", "StandardDev", "Success", "diff")

javelin_norm <- javelin_totals %>%
  ungroup() %>%
    mutate_at(aggstats, norm) %>%
      group_by(Athlete) %>%
        summarize_all(funs(mean))
```

&nbsp;  

### **4. Assigning weights to Athletes statistics**

To determine who the best athletes are, let's assign weights to the athletes for their distance, standard deviation of distance, if they were successful in their throw and the difference in their throws. Weights sum to 10. More weight placed in a statistic means it is more important.

```{r, results="hide"}
weights <- c(3.5, 2, 3, 1.5)

#creating total score by multiplying columns by their weights
javelin_team <- javelin_norm %>%
mutate(TotalScore = TotalDistance * weights[1] + StandardDev * weights[2] +
       Success + weights[3] + diff * weights[4]) %>%
arrange(desc(TotalScore))
```

&nbsp;  

### **5. Picking our team**

I'm going to take the top five athletes for our team from the dataframe we created by assigning the weights. Once we have our team, let's have a look at their stats.
```{r, results="hide"}

#selecting first five rows to pick athletes for our team
javelin_team<- javelin_team[1:5,]
javelin_team

#Selecting athlete names from our team
head(javelin_totals,1)
Ath <- javelin_team$Athlete

#Compiling our teams stats 
team_stats <- javelin_totals %>% 
  filter(Athlete %in% Ath) %>%
    summarize_all(funs(mean))
```
```{r, echo= FALSE}
pool_stats <- data.frame(do.call('cbind', sapply(javelin_totals, function(x) if(is.numeric(x)) c(max(x), mean(x)))))
pool_stats$MaxAve <- c("Maximum", "Average")
pool_stats <- pool_stats %>%
  gather(key="Statistic", value="Aggregate", -MaxAve)
```

And here's our team with their stats compared to that off the maximum and average of all athletes.
```{r, echo=FALSE}
(team_stats)
pool_stats
```

&nbsp;

### **6. How does my team stack up against the rest of the field?**

I'll see how my teams stats compares to the maximum and average of distance thrown as well as the maximum and average of the difference in an athletes last three and first three throws.
In a business setting, I could use these plots of their stats to justify why I would want these athletes on my team. We can see all our athletes throw well above the average, with two of our athletes throwing longer on their last three then first three throws. We could say these are our "clutch" players. 

The maximum stat is represented by the solid line and the average by the dashed.

```{r, echo=FALSE}
#A series of plots showing how each athlete compares to the maximum and the average of distance thrown and difference between their last three and first three throws
Max_d_line <- pool_stats[1,3]
Avg_d_line <- pool_stats[2,3]
Max_diff_line <- pool_stats[7,3]
Avg_diff_line <- pool_stats[8,3]

p <- team_stats %>%
  gather(Statistic, Aggregate, TotalDistance) %>%
  ggplot(aes(Athlete, Aggregate, fill = Athlete)) +
  geom_bar(stat = "identity", position = 'dodge') +
  geom_hline(yintercept =  Max_d_line, size = 2) +
  geom_hline(yintercept = Avg_d_line, linetype="dashed", size = 2) +
  labs(title=" My Team : Distance Thrown", color="Athlete pool maximum / average") +
  scale_fill_hue(l=70) +
  scale_color_hue(l=20) +
  theme_minimal() +
  theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank())

p

p_2 <- team_stats %>%
  gather(Statistic, Aggregate, diff) %>%
  ggplot(aes(Athlete, Aggregate, fill = Athlete)) +
  geom_bar(stat = "identity", position = 'dodge') +
  geom_hline(yintercept =  Max_diff_line, size = 2) +
  geom_hline(yintercept = Avg_diff_line, linetype="dashed", size = 2) +
  labs(title=" My Team : Difference between Last 3 and First 3 Throws", color="Athlete pool maximum / average") +
  scale_fill_hue(l=70) +
  scale_color_hue(l=20) +
  theme_minimal() +
  theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank())

p_2
```

### **7. Simulating a meet between my team and another**

Now, let's select  random athletes from athletes not on our team to play against us. We're going to create a fake meet to see if our teams total would be higher than the others.
```{r}
set.seed(77)
home <- c(2,3,5)
away <- sample(1:nrow(javelin_totals), 3, replace=FALSE)

HomeTeam <- round(sum(team_stats$TotalDistance[home]),2)
AwayTeam <- round(sum(javelin_totals$TotalDistance[away]),2)

print(paste0("Javelin match, Final Score: ", HomeTeam, " - ", AwayTeam))
ifelse(HomeTeam > AwayTeam, print("Your team won the match!"), print("Sorry, you lost."))
```

#### Our team won the match!
<br>